Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games

### 2025-06-07 research/paper/paper.tex
added data folder

\begin{abstract}
Multi-agent reinforcement learning has received increased attention in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms, such as independent Q-learning, have demonstrated simple implementation and versatility in various contexts. Previous research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we study the behaviour of these algorithms in non-cooperative settings. As a result, we explain the conditions that a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis. We collect empirical data on the performance of selected algorithms from experiments based on the proposed game. Our results show that independent value-based learning algorithms
can converge to Nash equilibrium in non-cooperative environments, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
%are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

\begin{keywords}
Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium
\end{keywords}

@@

\section{The Food Chain Game}
To verify our analysis, we test the algorithms in a game that can turn on and off the necessary condition given Equation~(\ref{eq:necessary_condition}) described in the previous section. We define the mechanics of the game Food Chain. An agent in the game represents an animal of some type. They can move around the map and feed themselves while trying to avoid enemies. Example game scenarios are shown in Figure~\ref{fig:biased_rps_in_food_chain} and Figure~\ref{fig:tree_maze_chase}. The game simulates an ecosystem to a certain extend, by defining a food web structure. A food web is a cluster of food chains showing what-eats-what in the ecosystem as illustrated in Figure~\ref{fig:food_chain_game}. The food web in the game is completely customizable and we can define an arbitrary number of species (agent types) and any number of agents of each type. This game is essentially a generalized version of the classic predator-and-prey problem. Recent research has tested IQL in the predator-and-prey game and found the algorithm to be seemingly performing well \cite{kopacz2023evaluating}. As we stated, these results could be misleading.

\begin{figure}
\centering
\subfloat[A Food Web]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_web.png}
    }
}
\subfloat[The Food Chain Game]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_chain_game.png}
    }
}
\caption{(a), A food web represents a simple ecosystem. The food web is a cluster of multiple food chains showing what-eats-what. The arrows point in the direction of the energy flow. (b), In the Food Chain game, an agent represents an individual animal of some type. They navigate on a playground trying to feed themselves while avoiding enemies.}
\label{fig:food_chain_game}
\end{figure}

>>>>

\begin{abstract}
Multi-agent reinforcement learning has received increased attention in cooperative games. However, research in non-cooperative games is lagging behind. Independent value-based learning algorithms have demonstrated simplicity and versatility in various contexts. In this paper, we study the behaviour of these algorithms in non-cooperative settings. We explain the conditions that a game must satisfy for the algorithms to work. We further test the algorithms in our proposed game Food Chain that simulates an ecosystem. Our results show that independent value-based learning algorithms can converge to Nash equilibrium, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
\end{abstract}

\begin{keywords}
Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium; food chain; food web; ecosystem
\end{keywords}

@@

\section{The Food Chain Game}
To verify our analysis, we test the algorithms in a game that can turn on and off the necessary condition given Equation~(\ref{eq:necessary_condition}) described in the previous section. We define the mechanics of the game Food Chain. An agent in the game represents an animal of some type. They can move around the map and feed themselves while trying to avoid enemies. The game simulates an ecosystem to a certain extend, by defining a food web structure. A food web is a cluster of food chains showing what-eats-what in the ecosystem as illustrated in Figure~\ref{fig:food_chain_game}. The food web in the game is completely customizable and we can define an arbitrary number of species (agent types) and any number of agents of each type. This game is essentially a generalized version of the classic predator-and-prey problem. Recent research has tested IQL in the predator-and-prey game and found the algorithm to be seemingly performing well \cite{kopacz2023evaluating}. As we stated, these results could be misleading.

\begin{figure}
\centering
\subfloat[A Food Web\textsuperscript{1}]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_web.png}
    }
}
\subfloat[The Food Chain Game\textsuperscript{1}]{
    \resizebox*{0.48\textwidth}{!}{
        \includegraphics{images/food_chain_game.png}
    }
}
\caption[Caption for LOF]{(a), A food web represents a simple ecosystem. The food web is a cluster of multiple food chains showing what-eats-what. The arrows point in the direction of the energy flow. (b), In the Food Chain game, an agent represents an individual animal of some type. They navigate on a playground trying to feed themselves while avoiding enemies.}
\label{fig:food_chain_game}
\end{figure}
\footnotetext[1]{The individual animal avatars were generated by ChatGPT GPT-4-turbo model which includes image generation via DALL·E 3, provided by OpenAI.}
